---
title: "Initial SOAR Monitoring Data Exploration"
author: "Jordanna Barley"
date: "11/9/2022"
output: word_document
---

```{r setup, include=FALSE}
require(here)
require(tidyverse)
require(readxl)
require(plotrix)

data<-read_excel(here('Data','MD','SOAR Monitoring Raw Data.xlsx'), sheet = 'DSOysData')
str(data)
data$Rep<- factor(data$Rep)
data$Substrate<-factor(data$Substrate)
data$OysSize<-factor(data$OysSize)
data$OysterStatus<- factor(data$OysterStatus)

ny22<-read_excel(here('Data','NY','BOP Governors Is SOAR Oysters.xlsx'), sheet = '2022 data')
ny21<-read_excel(here('Data','NY','BOP Governors Is SOAR Oysters.xlsx'), sheet ='pre-deployment')


ny21 %>%                                          
  ggplot(aes(x=bag_number, y=shell_height_mm))+
  geom_jitter(height = 1, width=0.3)
 
ny22 %>% 
  ggplot(aes(x=bag_number, y=shell_height_mm))+
  geom_jitter(height=1, width=0.3)

#attempting to calculate mean change in size
ny22_mean<-ny22 %>%                             #After SOAR mean size calculation by bag
  group_by(bag_number) %>% 
  summarise(mean_22=mean(shell_height_mm))

ny21_mean<-ny21%>%                              #before SOAR mean size calculation per bag
  group_by(bag_number) %>% 
  summarise(mean_21=mean(shell_height_mm))

ny_mean<-ny21_mean %>%                          #joining meean size data for both years
  inner_join(ny22_mean, by='bag_number') %>% 
  gather(year, mean, 2:3 ) %>% 
  mutate(year = paste(str_sub(year, -2,-1),bag_number))       #creating a new column that just has the year
  


ny22_sd<-ny22 %>%                               #After SOAR sd of size calculation per bag
  group_by(bag_number) %>% 
  summarise(sd_22=sd(shell_height_mm))

ny21_sd<-ny21 %>%                               #before SOAR sd of size calc per bag
  group_by(bag_number)%>% 
  summarise(sd_21=sd(shell_height_mm))

ny_sd<-ny22_sd %>%                              #joining the sd data for both years
  inner_join(ny21_sd, by='bag_number') %>% 
  gather(year, sd, 2:3) %>% 
  mutate(year = paste(str_sub(year, -2,-1),bag_number))

ny_size<-ny_sd %>%                              #joining the mean and sd data for both years
  full_join(ny_mean, by = c('year')) %>% 
  mutate(bag_number=bag_number.x, 
         year=substr(year, 1, 2)) %>% 
  select(bag_number, year, mean, sd) 
  
ny_size$year<-factor(ny_size$year)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

## What is in this document: 
Because we are still waiting for a fair amount of the data to come in, I picked sites where I knew we had sufficient data that would be good for a presentation. What I have here are very early plots and I would like some feeddback if this is what is needed for the upcoming Builder's meeting. 

### Maryland Monitoring Data
The Maryland monitoring was done by taking a sample of oysters from replicate quadrats, catagorizing them as live or box, and collecting size data. At present, I do not have pre-deployment data. I have computed some simple stats and produced some plots: 


```{r maryland, echo=FALSE, warning=FALSE}
data %>% 
  mutate(Rep=case_when(Rep=='001' ~ '1', 
                       Rep=='002' ~ '2',
                       Rep=='003' ~ '3', 
                       Rep=='004' ~ '4', 
                       Rep=='005' ~ '5', 
                       Rep=='006' ~ '6', 
                       Rep=='007' ~ '7', 
                       Rep=='008' ~ '8', 
                       Rep=='009' ~ '9', 
                       Rep=='010' ~ '10')) %>% 
  mutate(Rep = fct_relevel(Rep, '1','2','3','4','5','6','7','8','9','10'),
         OysterStatus = case_when(OysterStatus=='Box' ~ 'Dead',
                                  OysterStatus=='Live' ~ 'Live')) %>%  
  group_by(Rep) %>% 
  count(OysterStatus) %>% 
  ggplot(aes(x=Rep, y=n, fill=OysterStatus))+
  geom_bar(stat = 'identity', position='dodge')+
  theme_classic()+
  ylab('Count')+
  xlab('Replicate number')+
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size=18), 
        plot.title = element_text(size=22),
        legend.text = element_text(size=14),
        legend.title = element_text(size=14))

data %>% 
  mutate(Rep=case_when(Rep=='001' ~ '1', 
                       Rep=='002' ~ '2',
                       Rep=='003' ~ '3', 
                       Rep=='004' ~ '4', 
                       Rep=='005' ~ '5', 
                       Rep=='006' ~ '6', 
                       Rep=='007' ~ '7', 
                       Rep=='008' ~ '8', 
                       Rep=='009' ~ '9', 
                       Rep=='010' ~ '10')) %>% 
  mutate(Rep = fct_relevel(Rep, '1','2','3','4','5','6','7','8','9','10'), 
         OysterStatus = case_when(OysterStatus=='Box' ~ 'Dead',
                                  OysterStatus=='Live' ~ 'Live')) %>%  
  group_by(Rep) %>% 
  count(OysterStatus) %>% 
  group_by(Rep) %>% 
  mutate(prop = n/sum(n)) %>% 
  ggplot(aes(x=Rep, y=prop))+
  geom_point(size=2)+
  facet_wrap(~OysterStatus)+
  theme_bw()+
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size=18), 
        plot.title = element_text(size=22),
        legend.text = element_text(size=14),
        legend.title = element_text(size=14))+
  labs(x='Replicate Number',
       y='Proportion of total oyster sample')
  
data %>% 
  mutate(Rep=case_when(Rep=='001' ~ '1', 
                       Rep=='002' ~ '2',
                       Rep=='003' ~ '3', 
                       Rep=='004' ~ '4', 
                       Rep=='005' ~ '5', 
                       Rep=='006' ~ '6', 
                       Rep=='007' ~ '7', 
                       Rep=='008' ~ '8', 
                       Rep=='009' ~ '9', 
                       Rep=='010' ~ '10')) %>% 
  mutate(Rep = fct_relevel(Rep, '1','2','3','4','5','6','7','8','9','10')) %>%  
  group_by(Rep) %>% 
  count(OysterStatus) %>% 
  group_by(Rep) %>% 
  mutate(prop = n/sum(n)) %>% 
  filter(OysterStatus=='Live') %>% 
  ggplot(aes(x=Rep, y=prop))+
  geom_point(size=2.5)+
  theme_classic()+
  geom_hline(aes(yintercept=mean(prop)), color='red', size=1)+
  geom_hline(aes(yintercept=(mean(prop)+std.error(prop))), lty=2)+
  geom_hline(aes(yintercept=(mean(prop)-std.error(prop))), lty=2)+
  ylab('Proportion of live oysters')+
  xlab('Replicate number')+
  ggtitle('MD survival')+
  labs(caption = 'Proportion of live oysters collected from each replicate bag. Red line denotes the mean across replicates, dashed lines denote SEM.')+
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size=20),
        plot.caption = element_text(size=14)) 

data %>% 
  drop_na(Length) %>% 
  ggplot(aes(x=Length))+
  geom_histogram()+
  geom_vline(aes(xintercept = mean(Length)), color='blue', size=1)+
  ylab('Count')+
  xlab('Oyster Length (mm)')+
  labs(caption = 'Histogram of oyster lengths for all replicates. Blue line denotes the mean.')+
  theme_classic()+
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size=20), 
        plot.caption = element_text(size=14))
```

###New York Monitoring Data-Governor's Island
The data from this site were collected by taking a subset of 25 oysters from each replicate bag deployed in the water. We have size data for both pre- and post-deployment of the SOAR oysters from Govvernor's Island. The data do not really suggest significant growth across all replicates. I will compute growth, however I have not gotten there yet and wanted to share what I had so far. I can also compute some statistics to see if there is significant growth, though given the large standard deviations, there does not seem to be much difference in size between the two years. 

```{r New York data, echo=FALSE, warning=FALSE}
ny22 %>% 
  filter(sample_number=='1') %>% 
  mutate(prop_live=count_live/25) %>% 
  ggplot(aes(x=bag_number, y=prop_live))+
  geom_bar(stat='identity')+
  labs(x='Replicate Number', 
       y='Proportion of live oysters',
       title = 'Governors Island- Survival')+
  theme_classic()+
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size=20))

ny_size %>% 
  ggplot(aes(x=bag_number, y=mean, color=year))+
  geom_point(position=position_dodge(width=0.2,preserve = 'total'), size=2)+
  geom_linerange(aes(ymax=mean+sd, ymin=mean-sd), position=position_dodge(width = 0.2), size=1)+
  scale_color_manual(labels = c('Pre-deployment', 'Post-deployment'), values = c('#E69F00','#009E73'))+
  theme_classic()+
  ylab('Mean size (mm)')+
  xlab('Replicate number')+
  ggtitle('New York size data- Governors Island')+
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size=18), 
        plot.title = element_text(size=22),
        legend.text = element_text(size=14),
        legend.title = element_text(size=14))
```
